---
title: "R Notebook"
output: html_notebook
---




### Loading packages
```{r}
## For data wragnling & file management
library(tidyverse) 
library(stringr)
library(data.table)


## For analysis
library(rstatix)
library(spatstat)
library(vegan)

## For pretty plots
library(ggplot2)
library(ggpubr)
library(reshape2)
library(multcompView)
library(RColorBrewer)



set1_filePaths <- list.files(path = "Data/Set1 - Results/Phase 2", pattern = ".csv", full.names = TRUE)



```



## universal functions

```{r}

# Params finder: to stop the repeat reading in of files which takes ages
parse_filename <- function(file_name) {
  
  cndd <- str_extract(file_name, "cndd\\d+\\.?\\d*") %>%
    str_replace("cndd", "")
  
  hndd <- str_extract(file_name, "hndd\\d+\\.?\\d*") %>%
    str_replace("hndd", "")
  
  fragment_size <- str_extract(file_name, "s\\d+\\.?\\d*")
  
  fragment_no <- str_extract(file_name, "nf\\d+\\.?\\d*")

  immigration <- str_extract(file_name, "m\\d+\\.?\\d*")
  
  nf<- str_extract(file_name, "(?<=_nf)\\d+")
  
  L <- str_extract(file_name, "(?<=_s)\\d+")
  
  
  list(cndd = cndd, hndd = hndd, fragment_size = fragment_size, fragment_no = fragment_no, immigration = immigration, L = L, nf = nf )
}



```



# Set-1 Analysis

## Species diversity and distribution

### Shannons H

```{r}

findH <- function(file_name) {
  params <- parse_filename(file_name)
  cndd_value <- params$cndd
  hndd_value <- params$hndd
  
  df <- fread(file_name) ## >> read.csv
  
  shannon_results <- df %>%
    group_by(Repeat, TimeStep, Species) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = Species, values_from = n, values_fill = 0) ## very fast << always use over spread
  
  species_counts <- shannon_results %>% select(-Repeat, -TimeStep)
  
  shannon_results$H <- apply(species_counts, 1, shannonEntropy)
  avgH <- mean(shannon_results$H)
  
  tibble(hndd = hndd_value,
         cndd = cndd_value,
         avg_shannon_H = avgH)
}

shannon_raw_results <- lapply(set1_filePaths, findH)
shannon_summary <- bind_rows(shannon_raw_results)

shannon_summary$hndd <- factor(shannon_summary$hndd, levels = c("0", "0.5", "3"))
shannon_summary$cndd <- factor(shannon_summary$cndd, levels = c("0", "0.5", "3", "6", "9", "12"))

## Use this to store the results when needed
fwrite(shannon_summary, file = "Data/ShannonH_SET1.csv", row.names = FALSE)

```



### Species Abundances and Species Abundance Distributions

```{r}

calculate_sad <- function(file_path) {

  params <- parse_filename(file_path)
  cndd_value <- params$cndd
  hndd_value <- params$hndd

  data <- fread(file_path)
  
  ## Abundance calculations 
  species_abundance <- data %>% 
    group_by(Repeat, TimeStep, Species) %>% 
    summarise(abundance = n(), .groups = 'drop') %>% 
    arrange(Repeat, TimeStep, desc(abundance)) %>% 
    group_by(Repeat, TimeStep) %>% 
    mutate(rank = row_number()) %>% 
    ungroup()
  
  # mean abundance for each rank
  mean_sad <- species_abundance %>% 
    group_by(Repeat, TimeStep, rank) %>% 
    summarise(total_abundance = sum(abundance), .groups = 'drop') %>% 
    group_by(rank) %>% 
    summarise(mean_abundance = mean(total_abundance), .groups = 'drop')
  
  tibble(hndd = hndd_value, 
         cndd = cndd_value,
         rank = mean_sad$rank, 
         mean_abundance = mean_sad$mean_abundance)
}

sad_results <- lapply(set1_filePaths, calculate_sad)
sad_table <- bind_rows(sad_results)

sad_table$hndd <- factor(sad_table$hndd, levels = c("0", "0.5", "3"))
sad_table$cndd <- factor(sad_table$cndd, levels = c("0", "0.5", "3", "6", "9", "12"))

## you know what to do
fwrite(sad_table, file = "Data/SAD_SET1.csv", row.names = FALSE)
```


#### Plotting (Basic)

```{r}
sad_set1 <-  read_csv("Data/SAD_SET1.csv")
ShannonH_set1 <- read_csv("Data/ShannonH_SET1.csv")

sad_h_combined_set1 <- merge(sad_set1, ShannonH_set1, by = c("hndd", "cndd"))


sad_h_combined_set1$hndd <- factor(sad_h_combined_set1$hndd, levels = c("0", "0.5", "3"))
sad_h_combined_set1$cndd <- factor(sad_h_combined_set1$cndd, levels = c("12", "9", "6", "3", "0.5", "0"))

ggplot(sad_h_combined_set1)+
  geom_rect(aes(xmin = -Inf, xmax = Inf, 
                ymin = -Inf,ymax = Inf,
                fill = avg_shannon_H),
                alpha = 0.5, inherit.aes = FALSE)+
  geom_col(aes(x = rank, y = mean_abundance), colour = "red", width = 1) + 
  facet_grid(cndd ~ hndd)+ 
  #geom_blank(aes(x = rank, y = mean_abundance))+
  scale_fill_gradient(high = "#326b77", low = "#b5d1ae")+
  #scale_x_discrete(position = "top")+
  labs(x = "Strength of HNDD", y = "Stength of CNDD", fill = "", title = "Mean Shannon's Diversity Index (H)")+
  theme_pubr(legend = "right")+
  theme(
    text = element_text(family="serif"),
   # strip.background = element_blank(),
    strip.text = element_text(size = 15),
    axis.title.x = element_text(size = 15, vjust = 1),
    axis.title.y = element_text(size = 15),
    panel.spacing = unit(0, "lines"),
    panel.background = element_rect(fill = "grey", colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major =  element_blank(),
    panel.grid.minor =  element_blank(),
    axis.ticks = element_blank(),
    #panel.grid = element_line(linewidth = 1)
    axis.line.x = element_line(colour = "black"),
    panel.border = element_blank(),
    legend.key.height =  unit(3.5, "lines"),
    legend.title.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    legend.text = element_text(size = 12)
  )

```



## Spatial Statistics


### Clark Evans statistics

Crude measure of clustering vs ordering of a data set by comparing the average distances to the nearest neighbor vs the average distance to a random point (or nearest neighbor in a poisson distribution)


```{r}

## Warning very intensive function 

calculate_clark_evans <- function(file_path){
  
  params <- parse_filename(file_path)
  cndd_value <- params$cndd
  hndd_value <- params$hndd
  L <- params$L
  
  if(is.na(L)){
    L <- 1024 ## the default size for any unlabeled files 
  }
  
  
  data <- read_csv(file_path) ## Just making extra sure that I have a tidy data frame
  
  sizeWindow <- owin(c(0, L), c(0, L))
  
  clark_evans_results <- data %>% 
    group_by(Repeat, TimeStep, Species) %>% 
    do({ ## Do may be a little outdated but it gives a good progress bar and it makes sense in this case
      
      this_sp <- . 
      
      tpp <- ppp(x = this_sp$x, y = this_sp$y, window = sizeWindow) 
      
      CEval <-  clarkevans(tpp, correction = "Donnelly")
      
      tibble(CE = CEval, abundance= nrow(this_sp))
      
    }) %>% 
    filter(!is.infinite(CE)) 
  
  if(!is.data.frame(clark_evans_results)){
    stop("ERROR: outputting a non-data frame")
  }

  avg_ce <- mean(clark_evans_results$CE)

  tibble(
    hndd = factor(params$hndd),
    cndd = factor(params$cndd),
    mean_CE = avg_ce
  )
  
}

ce_results <- lapply(set1_filePaths, calculate_clark_evans)
ce_results <- bind_rows(ce_results)


ce_results$hndd <- factor(ce_results$hndd, levels = c("0", "0.5", "3"))
ce_results$cndd <- factor(ce_results$cndd, levels = c("0", "0.5", "3", "6", "9", "12"))

fwrite(ce_results, file ="Data/Clark_Evans_SET1.csv", row.names = FALSE)

```




#### Plotting
```{r}
CE_set1 <- read_csv("Data/Clark_Evans_SET1.csv")

CE_set1$hndd <- factor(CE_set1$hndd, levels = c("0", "0.5", "3"))
CE_set1$cndd <- factor(CE_set1$cndd, levels = c("0", "0.5", "3", "6", "9", "12"))

ggplot(CE_set1)+
  geom_tile(aes(x = hndd, y = cndd, fill = mean_CE)) + 
  scale_fill_gradient(high = "#326b77", low = "#b5d1ae")+
  labs(x = "Strength of HNDD", y = "Stength of CNDD", fill = "Mean Clark Evens Stat")+
  theme_minimal()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 10),
    panel.spacing = unit(0, "lines"),
    panel.background = element_rect(fill = "grey", colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major =  element_blank(),
    panel.grid.minor =  element_blank()
  )
```


### Pair correlation function / radial distibution function 

A more detailed understanding of how density varies as a function of distance from any one point. Here I am using it to understand the density of conspecifics from each other, ie. how far do plants disperse from their parent and still survive. (all closer together under the null as you would expect).


```{r}


calculate_pcf <- function(file_path) {
  
  params <- parse_filename(file_path)
  cndd_value <- params$cndd
  hndd_value <- params$hndd
  L <- params$L
  
   if(is.na(L)){
    L <- 1024 ## the default size for any unlabeled files 
  }
 
  data <- fread(file_path)
  
  window <- owin(c(0, L), c(0, L))
  
  ## sampling a subset of timesteps as this function is very computationally intensive 
  sampled_timesteps <- sample(x = unique(data$TimeStep), size = 20)
  
  
  data <- data %>% 
    filter(TimeStep %in% sampled_timesteps)
  
  # Filtering out species with < 20 individuals as they don't produce meaningful results (Can essentially be independent of each other)
  valid_species <- data %>%
    group_by(Repeat, TimeStep, Species) %>%
    summarise(count = n(), .groups = "drop") %>%
    filter(count >= 20)
  
  filtered_data <- data %>%
    semi_join(valid_species, by = c("Repeat", "TimeStep", "Species"))
  
  # Creating the radius and intervals for which to test the PCF function 
  R_seq <- seq(0, 200, by = 10)
  
  ## Running the PCF!
  pcf_results <- filtered_data %>%
    group_by(Repeat, TimeStep, Species) %>%
    group_modify(~ {
      tpp <- ppp(.x$x, .x$y, window = window)
      pcf_data <- pcf(tpp, correction = "Ripley", r = R_seq)
      
      tibble(
        r = pcf_data$r,
        iso = pcf_data$iso,
        abundance = nrow(.x)
      )
    }) %>%
    ungroup() %>%
    filter(!is.infinite(iso), !is.nan(iso))
  
 
  tibble(
    hndd = hndd_value,
    cndd = cndd_value,
    Repeat = as.factor(rep(filtered_data$Repeat[1], nrow(pcf_results))),
    TimeStep = as.factor(rep(filtered_data$TimeStep[1], nrow(pcf_results))),
    Species = rep(filtered_data$Species[1], nrow(pcf_results)),
    r = pcf_results$r,
    iso = pcf_results$iso,
    abundance = pcf_results$abundance
  )
}

pcf_results <- lapply(set1_filePaths, calculate_pcf)
pcf_table <- bind_rows(pcf_results)

pcf_table$hndd <- factor(pcf_table$hndd, levels = c("0", "0.5", "3"))
pcf_table$cndd <- factor(pcf_table$cndd, levels = c("0", "0.5", "3", "6", "9", "12"))

fwrite(pcf_table, file = "Data/PCF_SET1.csv", row.names = FALSE)

```


#### Plotting

```{r}
set1_pcf <- fread("Data/PCF_SET1.csv")


set1_pcf_summarised <- set1_pcf %>%
  filter(r <= 100, abundance > 50)

set1_pcf_summarised <- set1_pcf %>%
  group_by(hndd, cndd, r, Species) %>% 
  summarise(
    mean_iso = mean(iso),
    lower_ic = quantile(iso, 0.95), ## variability is important here
    upper_ic = quantile(iso, 0.05)
  )

ggplot() +
  geom_line(data = set1_pcf_summarised, aes(x = r, y = mean_iso))+
  geom_point(data = set1_pcf_summarised, aes(x = r, y = mean_iso))+
  geom_ribbon(data = set1_pcf_summarised, aes(x = r, ymin = lower_ic, ymax = upper_ic), alpha = 0.3)+ 
  facet_grid(~cndd~hndd)+
  theme_minimal()
```




















